Interpretable Neural Networks using EAGGA

### 2025-03-14 paper/main.tex
HPO model-free

\subsection{Hyperparameter Optimization (HPO)}
In contrast to model parameters, which are optimized during training, hyperparameters are those describing the machine learning algorithm and are fixed before training.
Because they usually have a significant impact on the trained model's performance, hyperparameters are often optimized, too.
Let $\mathcal{D}\subseteq\mathcal{X}\times\mathcal{Y}$ be a dataset consisting of $n$
tuples drawn from the data-generating distribution, i.e. $(\boldsymbol{x}^{(i)}, y^{(i)})\stackrel{i.i.d.}{\sim}\mathbb{P}_{\boldsymbol{x}y},\forall i=1,...,n$.
Furthermore, let $\mathcal{I}:(\mathbb{D}\times\boldsymbol\Lambda)\rightarrow\mathcal{H}, (\mathcal{D},\boldsymbol\lambda)\mapsto\hat{f}$
be an algorithm that maps a given dataset $\mathcal{D}$ and hyperparameter configuration $\boldsymbol\lambda\in\boldsymbol\Lambda$ to a model $\hat{f}$.
Denote with $\mathcal{I}_{\boldsymbol\lambda}$ an algorithm with fixed hyperparameter configuration a chosen loss function with $L$.
The goal of model training is, for fixed $\boldsymbol\lambda$, to have $\mathcal{I}_{\boldsymbol\lambda}$ find model $\hat{f}$ minimising the expected generalisation error
$GE(\mathcal{I}_{\boldsymbol\lambda},\mathcal{D},L)=\mathbb{E}_{(\boldsymbol{x},y)\sim\mathbb{P}_{\boldsymbol{x}y}}[L(y,\mathcal{I}_{\boldsymbol\lambda}(\mathcal{D})(\boldsymbol{x}))]$
The goal of HPO, on the other hand, is to find $\boldsymbol\lambda$ minimising the expected generalisation error,
i.e. $\argmin_{\boldsymbol\lambda\in\boldsymbol\Lambda} GE(\mathcal{I}_{\boldsymbol\lambda},\mathcal{D},L)$.
For this there is no analytical expression available, making HPO a blackbox optimization problem.  \citep[pp. 2f]{10.1145/3610536}

[model free]
(grid + random search)
- most basic strategy: grid search, for all HPs define range of interest, then evaluate cartesian product of those -> scales poorly in \# of HP dimensions + \# of query points per HP range
- random search
  * randomly sample value for each HP until it runs out of budget
  * explorative: for budget $B$, each HP will (likely) be queried with $B$ different values, whereas grid search only $B^{1/N}$ for $N$ HPs
  * thus better than grid in case of some HPs being non-informative (almost always the case), as it doesn't waste time specifically exploring these (as does grid)
  * usually useful as baseline, as no assumptions about model, easy to parallelise, high exploration, and in expectation will ``achieve performance arbitrarily close to the optimum''

>>>>

\subsection{Hyperparameter Optimization (HPO)}
In contrast to model parameters, which are optimized during training, hyperparameters (HPs) are those describing the machine learning algorithm and are fixed before training.
Because they usually have a significant impact on the trained model's performance, HPs are often optimized, too.
Let $\mathcal{D}\subseteq\mathcal{X}\times\mathcal{Y}$ be a dataset consisting of $n$
tuples drawn from the data-generating distribution, i.e. $(\boldsymbol{x}^{(i)}, y^{(i)})\stackrel{i.i.d.}{\sim}\mathbb{P}_{\boldsymbol{x}y},\forall i=1,...,n$.
Furthermore, let $\mathcal{I}:(\mathbb{D}\times\boldsymbol\Lambda)\rightarrow\mathcal{H}, (\mathcal{D},\boldsymbol\lambda)\mapsto\hat{f}$
be an algorithm that maps a given dataset $\mathcal{D}$ and HP configuration $\boldsymbol\lambda\in\boldsymbol\Lambda$ to a model $\hat{f}$.
Denote with $\mathcal{I}_{\boldsymbol\lambda}$ an algorithm with fixed HP configuration a chosen loss function with $L$.
The goal of model training is, for fixed $\boldsymbol\lambda$, to have $\mathcal{I}_{\boldsymbol\lambda}$ find model $\hat{f}$ minimising the expected generalisation error
$GE(\mathcal{I}_{\boldsymbol\lambda},\mathcal{D},L)=\mathbb{E}_{(\boldsymbol{x},y)\sim\mathbb{P}_{\boldsymbol{x}y}}[L(y,\mathcal{I}_{\boldsymbol\lambda}(\mathcal{D})(\boldsymbol{x}))]$
The goal of HPO, on the other hand, is to find $\boldsymbol\lambda$ minimising the expected generalisation error,
i.e. $\argmin_{\boldsymbol\lambda\in\boldsymbol\Lambda} GE(\mathcal{I}_{\boldsymbol\lambda},\mathcal{D},L)$.
For this there is no analytical expression available, making HPO a blackbox optimization problem. \citep[pp. 2f]{10.1145/3610536}

We will first outline model-free methods to approach blackbox optimization problems, before presenting a model-based variant.

% model free
% grid + random search
The most basic model-free strategies for blackbox optimization are grid and random search.
In the latter, the user defines a range of interest for each HP and then evaluating points along a grid within the cartesian product of those.
This has the drawback of scaling extremely poorly in both number of HP dimensions and number of query points per range of interest.
Random search, on the other hand, randomly samples a value for each HP until it runs out of budget. Its explorative nature, ease of use, and no assumptions about the model
makes it a good baseline. It is also superior to grid search in cases were one or more HPs have little impact on model performance, as for a given budget $B$,
each HP will likely be queried with $B$ different values, whereas for grid search each HP will only be queried with $B^{1/N}$ different values for $N$ HPs.
